Do Alternate Wetting and Drying Irrigation Technology and Nitrogen Rates Affect Rice Sheath Blight?

### 2018-11-02 analysis/paper/paper.Rmd
Wordsmithing and other minor edits

---
title: "Do Alternate Wetting and Drying Irrigation Technology and Nitrogen Rates Affect Rice Sheath Blight?"
author: |
  | Adam H. Sparks^[Corresponding author: Adam.Sparks@usq.edu.au]  \textsuperscript{1\dag}, Nancy P. Castilla \textsuperscript{1}, Bjorn Ole Sander \textsuperscript{1}
  | \textsuperscript{1}International Rice Research Institute, Crop and Environmental Sciences Division, Los Baños, Laguna, Philippines, 4031
  | \textsuperscript{\dag} Current Address: University of Southern Queensland, Centre for Crop Health, West Street, Tooowoomba, Queensland, Australia, 4350
csl: "../templates/european-journal-of-plant-pathology.csl" # Insert path for the bib-style
abstract: |
  Water and nitrogen management play vital roles in rice production. However, the mismanagement of these two management practices may trigger sheath blight of rice, caused by _Rhizoctonia solani_, which is favored by wet conditions, high relative humidity, and high nitrogen fertilizer levels. To understand how different combinations of water and nitrogen management affect sheath blight epidemics, we conducted two separate split--plot experiments with a water saving (alternate wetting and drying) regime and a traditional puddled regime combined with differing nitrogen treatments in the dry seasons of 2015 and 2016. Disease was scored in the same way in both experiments using a sheath blight assessment scale for field evaluation developed at the International Rice Research Institute to assess the severity on infected sheaths and leaves while sheath blight incidence on tillers were counted per hill. We were unable to detect any difference in the incidence of tiller sheath blight due to irrigation, tiller and leaf sheath blight did clearly differ statistically by irrigation treatment, but leaf sheath blight severity did not. Our findings suggest that farmers can adopt water saving technologies without risking increased sheath blight incidence. We suggest that further cross--disciplinary research in this area is warranted. All data generated from this project and code used for analysis are available as a reproducible research compendium from <https://github.com/openplantpathology/rice_awd_pests> [@Sparks2018].
bibliography: references.bib
output:
  bookdown::word_document2:
    toc: no
    reference_docx: "../templates/template.docx"
---

@@

Alternate wetting and drying (AWD) is an irrigation technique for rice (*Oryza sativa* L.) developed by the International Rice Research Institute (IRRI) and its partners that saves about 15--40&nbsp;\% of irrigation water while not affecting yield [@Bouman2001; @Feng2007]. In AWD rice, fields are exposed to several dry phases during the growing season without exposing the plants to water stress. In order to avoid yield decline under AWD “safe” thresholds have been developed. Under safe AWD irrigation water is applied when the field water level reaches 15&nbsp;cm below the soil surface [@Richards2014]. Fields are furthermore kept puddled during the flowering period to avoid plant stress. As an added benefit to saving water, AWD also reduces greenhouse gas (GHG) emissions of rice fields, which are a substantial factor in the GHG budget of rice producing countries, by around 50 &nbsp;\% [@Yan2005; @Sander2015].

@@

The 2016 experiment followed the same split--plot design as described for the 2015 experiment. However, the plot size was increased and due to these changes, the replicate sizes are are not equal as necessitated by the use of a larger area for the experiment. The main plot sizes were: Block 1 (B1) 21&nbsp;m x 20.5&nbsp;m (412.5 sq&nbsp;m) and Block 2 (B2) 20.25&nbsp;m x 21.6&nbsp;m (437.4&nbsp;m^2^). The split--plot sizes were B1 21&nbsp;m x 10.25&nbsp;m (215.25&nbsp;m^2^), B2 20.25&nbsp;m x 10.8&nbsp;m (218.7&nbsp;m^2^). The replication sizes were B1 --- 42&nbsp;m&nbsp;x&nbsp;20.5&nbsp;m (861&nbsp;m^2^) and B2 --- 40.5&nbsp;m&nbsp;x&nbsp;21.6&nbsp;m (874.8&nbsp;m^2^). A buffer 0.5&nbsp;m per split--plot was used and the overall experiment size was 3471.6&nbsp;m^2^.

split--plot N rates differed from the 2015 rates with only two N treatments, N60 (60&nbsp;kg&nbsp;/&nbsp;ha as urea split into two applications) and N180 (180&nbsp;kg&nbsp;/&nbsp;ha in three splits) being applied (Tab.&nbsp;\@ref(tab:N_rate)).

@@

Based on the 2015 results, the inoculation methods were modified in 2016 to increase the amount of inoculum and applied to a smaller area. Plots were inoculated 41 days after transplanting using ten bottles per one sampling area (1&nbsp;m&nbsp;x&nbsp;1&nbsp;m) per plot, where one bottle contained 151&nbsp;g inoculum. A total amount of 1,510&nbsp;g of inoculum was applied to a 1&nbsp;m&nbsp;x&nbsp;1&nbsp;m area.

@@

Area under the disease progress stairs (AUDPS) [@simko2012area] was calculated for both disease incidence and severity using the R package `agricolae` [@R2018; @deMendiburu2017]. Due to the scale used for collection, the severity data were converted to the midpoint percent value of the scale range for each value.

@@

As most of the data's residuals did not meet assumptions for normality, the analysis was carried out using Bayesian multivariate generalised linear mixed models implemented in the R package `MCMCglmm` [@MCMCglmm2010]. Six models were created where the the dependent variables were 2015 tiller sheath blight incidence, 2015 tiller sheath blight severity, 2015 leaf sheath blight severity; 2016 tiller sheath blight incidence, 2016 tiller sheath blight severity; 2016 leaf sheath blight severity, respectively. The independent variables were the irrigation management and N rate with replicate treated as a random variable. The base levels (control treatments) for the analyses for 2015 were: N rate -- N0, irrigation management -- puddled and for 2016 they were: N rate -- N60, irrigation management -- puddled

@@

In 2015 the incidence of tiller sheath blight remained low throughout the growing season (Fig.&nbsp;\@ref(fig:fig-1)a, \@ref(fig:fig-2)a). Water management was not clearly statistically different (Fig.&nbsp;\@ref(fig:fig-3)a, \@ref(fig:fig-5)a). However, the N treatments, N100 and N120, were both clearly statistically different when compared with the control N0 treatment (Fig.&nbsp;\@ref(fig:fig-4)a, \@ref(fig:fig-5)a).

@@

Tiller sheath blight severity remained below 2&nbsp;\% (Fig.&nbsp;\@ref(fig:fig-2)c, \@ref(fig:fig-1)c). Both the N100 and N120 treatments were clearly statistically different than the control N0 treatment (Fig.&nbsp;\@ref(fig:fig-4)c, \@ref(fig:fig-5)c). However, the AWD water management was not clearly statistically different from the puddled treatment (Fig.&nbsp;\@ref(fig:fig-3)c, \@ref(fig:fig-5)c).

@@

Severity of leaf sheath blight remained low, less than 0.4&nbsp;% across all treatments (Fig.&nbsp;\@ref(fig:fig-1)e, \@ref(fig:fig-2)e). Both the N100 and N120 treatments were clearly statistically different than the N0 treatment (Fig.&nbsp;\@ref(fig:fig-4)e, \@ref(fig:fig-5)e). However, the AWD water management was not clearly statistically different from the puddled treatment (Fig.&nbsp;\@ref(fig:fig-3)e, \@ref(fig:fig-5)e).

@@

The changes to the inoculation methodology resulted in a higher rate of infection in 2016, with the N180 treatment reaching a maximum value of 98&nbsp;% incidence at the third observation (Fig.&nbsp;\@ref(fig:fig-2)b). The N treatment N180, was clearly statistically different than the control N60 treatment (Fig.&nbsp;\@ref(fig:fig-4)b, \@ref(fig:fig-5)b). As in the 2015 study, water management did not clearly statistically differ (Fig.&nbsp;\@ref(fig:fig-3)b, \@ref(fig:fig-5)b).

@@

As with the tiller incidence, the tiller severity increased with the changed inoculation methods with a maximum of 7.6% for the puddled treatment (Fig.&nbsp;\@ref(fig:fig-2)d). In 2016 the N180 treatment was clearly statistically different from the N60 treatment, with N180 severity being higher (Fig.&nbsp;\@ref(fig:fig-4)d, \@ref(fig:fig-5)d). The AWD water management, which was lower than the puddled treatment, was also clearly statistically different (Fig.&nbsp;\@ref(fig:fig-3)d, \@ref(fig:fig-5)d).

@@

In 2016 the neither of the treatments, N rate or water management, were clearly statistically different from the control treatments for leaf sheath blight severity (Fig.&nbsp;\@ref(fig:fig-3)f, \@ref(fig:fig-5)f, \@ref(fig:fig-4)f, \@ref(fig:fig-5)f).

@@

By increasing the plot size and increasing inoculum amount applied to a smaller area, the changes made for the 2016 experiment appear to have improved the experiment. The the sheath blight incidence increased and the variability of sheath blight in the plots decreased (Fig.&nbsp;\@ref(fig:fig-1)a:b, \@ref(fig:fig-2)c:d, \@ref(fig:fig-3)a:b, \@ref(fig:fig-4)c:d).

It should be noted that the levels of leaf severity remained low (< 1 \%) throughout the growing season in all treatments for both years. As AWD can only feasibly be implemented in the dry season, the sheath blight disease levels in these studies may be lower than they would be if conducted in the wet season. However, sheath blight remains an issue in both seasons and the use of AWD would not appear to increase the disease but may be effective in reducing it in some cases.

The rice establishment method is known to affect the spread of sheath blight [@Willocquet2000]. Manually transplanted rice had higher apparent infection rates than direct seeded rice. As manual transplanting becomes less common due to labor constraints, sheath blight risk may be decreased.

@@

The authors wish to thank the IRRI Biometrics group for their input in the design of the experiments and Rachel King of the University of Southern Queensland Statistical Consulting Unit for her insightful comments on the analysis of this paper. We would also like to thank Dale Amozola, Paul Escandor and P. Lapis for their technical support in the field. Lastly, we would like to remember our team mate Michael Noel, a part of the plant disease management team at IRRI and who helped design and managed the first year of the project but passed away in September 2016.

>>>>

---
title: "Do Alternate Wetting and Drying Irrigation Technology and Nitrogen Rates Affect Rice Sheath Blight?"
author: |
  | Adam H. Sparks^[Corresponding author: Adam.Sparks@usq.edu.au]  \textsuperscript{1\dag}, Nancy P. Castilla \textsuperscript{1}, Bjorn Ole Sander \textsuperscript{1}
  | \textsuperscript{1}International Rice Research Institute, Crop and Environmental Sciences Division, Los Baños, Laguna, Philippines, 4031
  | \textsuperscript{\dag} Current Address: University of Southern Queensland, Centre for Crop Health, West Street, Tooowoomba, Queensland, Australia, 4350
csl: "../templates/european-journal-of-plant-pathology.csl" # Insert path for the bib-style
abstract: |
  Water and nitrogen management play vital roles in rice production. However, the mismanagement of these two management practices may trigger sheath blight of rice, caused by _Rhizoctonia solani_, which is favored by wet conditions, high relative humidity, and high nitrogen fertilizer levels. To understand how different combinations of water and nitrogen management affect sheath blight epidemics, we conducted two separate split--plot experiments with a water saving (alternate wetting and drying) regime and a traditional puddled regime combined with differing nitrogen treatments in the dry seasons of 2015 and 2016. Disease was scored in the same way in both experiments using a sheath blight assessment scale for field evaluation developed at the International Rice Research Institute to assess the severity on infected sheaths and leaves while sheath blight incidence on tillers were counted per hill. We were unable to detect any difference in the incidence of tiller sheath blight due to irrigation, tiller and leaf sheath blight did clearly differ statistically by irrigation treatment, but leaf sheath blight severity did not. Our findings indicate that farmers can adopt water saving technologies without risking increased sheath blight incidence. We suggest that further cross--disciplinary research in this area is warranted. All data generated from this project and code used for analysis are available as a reproducible research compendium from <https://github.com/openplantpathology/rice_awd_pests> [@Sparks2018].
bibliography: references.bib
output:
  bookdown::word_document2:
    toc: no
    reference_docx: "../templates/template.docx"
---

@@

Alternate wetting and drying (AWD) is an irrigation technique for rice (*Oryza sativa* L.) developed by the International Rice Research Institute (IRRI) and its partners that saves about 15--40&nbsp;\% of irrigation water while not affecting yield [@Bouman2001; @Feng2007]. In AWD rice, fields are exposed to several dry phases during the growing season without exposing the plants to water stress. In order to avoid yield decline under AWD "safe"" thresholds have been developed. Under safe AWD irrigation water is applied when the field water level reaches 15&nbsp;cm below the soil surface [@Richards2014]. Fields are furthermore kept puddled during the flowering period to avoid plant stress. As an added benefit to saving water, AWD also reduces greenhouse gas (GHG) emissions of rice fields, which are a substantial factor in the GHG budget of rice producing countries, by around 50 &nbsp;\% [@Yan2005; @Sander2015].

@@

The 2016 experiment followed the same split--plot design as described for the 2015 experiment. However, the plot size was increased and due to these changes, the replicate sizes are are not equal as necessitated by the use of a larger area for the experiment due to differences in the blocks. The main plot sizes were: Block 1 (B1) 21&nbsp;m x 20.5&nbsp;m (412.5 sq&nbsp;m) and Block 2 (B2) 20.25&nbsp;m x 21.6&nbsp;m (437.4&nbsp;m^2^). The split--plot sizes were B1 21&nbsp;m x 10.25&nbsp;m (215.25&nbsp;m^2^), B2 20.25&nbsp;m x 10.8&nbsp;m (218.7&nbsp;m^2^). The replication sizes were B1 --- 42&nbsp;m&nbsp;x&nbsp;20.5&nbsp;m (861&nbsp;m^2^) and B2 --- 40.5&nbsp;m&nbsp;x&nbsp;21.6&nbsp;m (874.8&nbsp;m^2^). A buffer 0.5&nbsp;m per split--plot was used and the overall experiment size was 3471.6&nbsp;m^2^.

Split--plot N rates differed from the 2015 rates with only two N treatments, N60 (60&nbsp;kg&nbsp;/&nbsp;ha as urea split into two applications) and N180 (180&nbsp;kg&nbsp;/&nbsp;ha in three splits) being applied (Tab.&nbsp;\@ref(tab:N_rate)).

@@

Based on the 2015 results, the inoculation methods were modified in 2016 to increase the amount of inoculum and applied to a smaller area. Plots were inoculated 41 days after transplanting using ten bottles per one sampling area (1&nbsp;m&nbsp;x&nbsp;1&nbsp;m) per plot, where one bottle contained 151&nbsp;g inoculum. A total amount of 1,510&nbsp;g of inoculum was applied to each 1&nbsp;m&nbsp;x&nbsp;1&nbsp;m area.

@@

Area under the disease progress stairs (AUDPS) [@simko2012area] was calculated for both disease incidence and severity using R, version 3.5.0 [@R2018], package `agricolae`, version 1.2-8, [@deMendiburu2017]. Due to the use of a severity scale for data collection, the severity data were converted to the midpoint percent value of the scale range for each value.

@@

As most of the data's residuals did not meet assumptions for normality, the analysis was carried out using Bayesian multivariate generalised linear mixed models implemented in the R package `MCMCglmm` [@MCMCglmm2010], version 2.26. Six models were created where the the dependent variables were 2015 tiller sheath blight incidence, 2015 tiller sheath blight severity, 2015 leaf sheath blight severity; 2016 tiller sheath blight incidence, 2016 tiller sheath blight severity; 2016 leaf sheath blight severity, respectively. The independent variables were the irrigation management and N rate with replicate treated as a random variable. The base levels (control treatments) for the analyses for 2015 were: N rate -- N0, irrigation management -- puddled and for 2016 they were: N rate -- N60, irrigation management -- puddled

@@

In 2015 the incidence of tiller sheath blight remained low throughout the growing season (Fig.&nbsp;\@ref(fig:fig-1)a, \@ref(fig:fig-2)a). Water management was not clearly statistically different (Fig.&nbsp;\@ref(fig:fig-3)a, \@ref(fig:fig-5)a). However, the effects of N treatments N100 and N120 on tiller incidence caused both to be higher and clearly statistically different when compared with the control N0 treatment (Fig.&nbsp;\@ref(fig:fig-4)a, \@ref(fig:fig-5)a).

@@

Tiller sheath blight severity remained below 2&nbsp;\% (Fig.&nbsp;\@ref(fig:fig-2)c, \@ref(fig:fig-1)c). Both the N100 and N120 treatment effects caused the tiller severity to be higher and clearly statistically different than the control N0 treatment (Fig.&nbsp;\@ref(fig:fig-4)c, \@ref(fig:fig-5)c). However, the AWD water management was not clearly statistically different from the puddled treatment (Fig.&nbsp;\@ref(fig:fig-3)c, \@ref(fig:fig-5)c).

@@

Severity of leaf sheath blight remained low, less than 0.4&nbsp;% across all treatments (Fig.&nbsp;\@ref(fig:fig-1)e, \@ref(fig:fig-2)e). As with the tiller incidence and severity, the effects of both the N100 and N120 treatments on leaf severity resulted in higher values and were and clearly statistically different than the N0 treatment (Fig.&nbsp;\@ref(fig:fig-4)e, \@ref(fig:fig-5)e). However, the AWD water management was not clearly statistically different from the puddled treatment (Fig.&nbsp;\@ref(fig:fig-3)e, \@ref(fig:fig-5)e).

@@

The changes to the inoculation methodology resulted in a higher rate of infection in 2016, with the N180 treatment reaching a maximum value of 98&nbsp;% incidence at the third observation (Fig.&nbsp;\@ref(fig:fig-2)b). The N treatment N180 value was higher and clearly statistically different than the control N60 treatment (Fig.&nbsp;\@ref(fig:fig-4)b, \@ref(fig:fig-5)b). As in the 2015 study, water management did not clearly statistically differ (Fig.&nbsp;\@ref(fig:fig-3)b, \@ref(fig:fig-5)b).

@@

As with the tiller incidence, the tiller severity increased with the changed inoculation methods with a maximum of 7.6% for the puddled treatment (Fig.&nbsp;\@ref(fig:fig-2)d). In 2016 the N180 treatment was again higher and clearly statistically different from the N60 treatment, with N180 severity being higher (Fig.&nbsp;\@ref(fig:fig-4)d, \@ref(fig:fig-5)d). The AWD water management, which was lower than the puddled treatment, was also clearly statistically different (Fig.&nbsp;\@ref(fig:fig-3)d, \@ref(fig:fig-5)d).

@@

In 2016 the neither of the treatments, N rate or water management, were clearly statistically different from the control treatments, N60 and puddling water management, for leaf sheath blight severity (Fig.&nbsp;\@ref(fig:fig-3)f, \@ref(fig:fig-5)f, \@ref(fig:fig-4)f, \@ref(fig:fig-5)f).

@@

By increasing the plot size and increasing inoculum amount applied to a smaller area, the changes made for the 2016 experiment appear to have improved the experiment. The sheath blight incidence increased and the variability of sheath blight in the plots decreased (Fig.&nbsp;\@ref(fig:fig-1)a:b, \@ref(fig:fig-2)c:d, \@ref(fig:fig-3)a:b, \@ref(fig:fig-4)c:d). Based on these results, it is suggested that the methods used in the 2016 experiment should be adopted in any further studies of this kind.

It should be noted that the levels of leaf severity remained low (< 1 \%) throughout the growing season in all treatments for both years. As AWD can only feasibly be implemented in the dry season. Therefor, the sheath blight disease levels in these studies may be lower than they would be if conducted in the wet season. However, sheath blight remains an issue in both seasons and the use of AWD would not appear to increase the disease but may be effective in reducing it in some cases.

As the rice establishment method is known to affect the spread of sheath blight [@Willocquet2000]. Manually transplanted rice had higher apparent infection rates than direct seeded rice. As manual transplanting becomes less common due to labor constraints, sheath blight risk may be decreased (CITATION XXXX) due changes in establishment practices. A combination of AWD and directd seeded rice could be beneficial for both saving water and reducing the risk of sheath blight.

@@

The authors wish to thank the IRRI Biometrics group for their input in the design of the experiments and Dr Rachel King of the University of Southern Queensland Statistical Consulting Unit for her insightful comments on the analysis of this paper. We would also like to thank Dale Amozola, Paul Escandor and P. Lapis for their technical support in the field. Lastly, we would like to remember our team mate Michael Noel, a part of the plant disease management team at IRRI and who helped design and managed the first year of the project but passed away in September 2016.
